---
title: "MB Habituation - Habituation Curve"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(here)
library(zoo)

theme_set(theme_bw() + 
            theme(panel.grid = element_blank()))
```

## Load preprocessed data

```{r load}
d <- read_csv(here("data","processed_data","final_merged_manybabies_preprocessed.csv"))
  
d |> count(project_id, name = "n_trials")
```

```{r}
d_proc <- d |> 
  mutate(
    project_id = project_id |> factor() |> `contrasts<-`(value = contr.sum(3) * 0.5),
    looking_time_corr = ifelse(
      project_id == "MB4" & trial_type == "social",
      pmax(0.01, (looking_time - 4.4)),
      looking_time
    )) |> 
  group_by(unique_subject_identifier) |> 
  mutate(lt_mov_avg = rollmean(looking_time, 3, na.pad = T),
         lt_baseline = lt_mov_avg[2],
         lt_habituated = lt_mov_avg < 0.5 * lt_baseline)
```

## Visual check

```{r plot, fig.width=8, fig.height=5}
d |>
  group_by(project_id, trial_num) |>
  summarise(mean_lt = mean(looking_time), .groups = "drop") |>
  ggplot(aes(trial_num, mean_lt, colour = project_id)) +
  geom_point() +
  geom_smooth(aes(linetype = "Linear"),    method = "lm",
              formula = y ~ x,             se = FALSE) +
  geom_smooth(aes(linetype = "Quadratic"), method = "lm",
              formula = y ~ poly(x, 2),    se = FALSE) +
  labs(x = "Trial number", y = "Mean looking time (s)",
       title = "Habituation: linear vs. quadratic", linetype = "Form") +
  theme_minimal()
```

```{r plot, fig.width=8, fig.height=5}
d |>
  group_by(project_id, trial_num) |>
  summarise(mean_log_lt = mean(log_lt), .groups = "drop") |>
  ggplot(aes(trial_num, mean_log_lt, colour = project_id)) +
  geom_point() +
  geom_smooth(aes(linetype = "Linear"),    method = "lm",
              formula = y ~ x,             se = FALSE) +
  geom_smooth(aes(linetype = "Quadratic"), method = "lm",
              formula = y ~ poly(x, 2),    se = FALSE) +
  labs(x = "Trial number", y = "Mean log looking time (s)",
       title = "Habituation: linear vs. quadratic", linetype = "Form") +
  theme_minimal()
```

## Moving average habituation criterion

```{r}
d_habituated <- d_proc |> 
  filter(lt_habituated) |> 
  slice(1)

d_habituated |> 
  ggplot(aes(x = trial_num, fill = project_id)) +
  geom_histogram(binwidth = 1)
```

Very few kids in MB4 habituated, which is not very surprising (many of them maxed out the looking time).

## MB4 DV exploration
MB4 has two DVs: the LT on the video itself, and the LT on the freeze frame at the end of the video. 
Furthermore, the videos differed across conditions, with the social condition video being 4.4s longer than the nonsocial condition video. 

We explore the goodness of different DV choices (LT_video, LT_frame, LT_video_frame, LT_video_corr, LT_video_frame_corr), as well as LT_video with an additional fixed effect of trial_type.
One option is to fit the power law fit and see which one has the best fit.

```{r}
mb4_formula_list <- list(
  formula(log(looking_time + .01) ~ log(trial_num) + (1 + trial_num | lab_id / subject)),
  formula(log(lookingtime_freezeframe + .01) ~ log(trial_num) + (1 + trial_num | lab_id / subject)),
  formula(log(looking_time + lookingtime_freezeframe + .01) ~ log(trial_num) + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time_corr + .01) ~ log(trial_num) + (1 + trial_num | lab_id / subject)),
  formula(log(looking_time_corr + lookingtime_freezeframe + .01) ~ log(trial_num) + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time + .01) ~ log(trial_num) * trial_type + (1 + trial_num | lab_id / subject))
)

mb4_models <- tibble(
  formula = mb4_formula_list,
  model = map(formula, \(f) {
    lmer(f, data = d_proc)
  }),
  BIC = map_dbl(model, BIC)
)

mb4_models |> select(-model)
```

Observations:
- All models with freezeframe LT did not converge
- The best fitting model was the model with raw LT and trial type

## Models
```{r}
formula_list <- list(
  formula(looking_time ~ trial_num * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ poly(trial_num, 2) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ log(trial_num) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ trial_num * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ poly(trial_num, 2) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ log(trial_num) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ trial_num * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ poly(trial_num, 2) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ log(trial_num) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ trial_num * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ poly(trial_num, 2) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ log(trial_num) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject))
)
```

```{r}
models <- tibble(
  lt = rep(c("lt", "log_lt"), times = 2, each = 3),
  fxn = rep(c("linear", "quadratic", "log"), times = 4),
  cond = rep(c("no", "yes"), each = 6),
  formula = formula_list,
  model = map(formula, \(f) {
    lmer(f, data = d_proc)
  }),
  BIC = map_dbl(model, BIC)
)

models |> select(-formula, -model) |> arrange(BIC)
```

```{r}
d_preds <- d_proc |> 
  ungroup() |> 
  mutate(pred = predict(models$model[[12]]))
```

## Some plots

```{r}
d_proc |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = looking_time_corr, fill = trial_type)) +
  geom_histogram()  
```

```{r}
d_preds |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = looking_time, y = lookingtime_freezeframe, col = lab_id, shape = trial_type)) +
  geom_hline(yintercept = 12) +
  geom_point(alpha = .3)
```

```{r}
d_preds |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = trial_num, y = lookingtime_freezeframe, col = trial_type)) +
  geom_point(alpha = .2, position = position_jitter(width = .2)) +
  geom_smooth(alpha = .1)
```

## Quantifying each child's slope
```{r}
power_mod <- models |> 
  filter(lt == "log_lt", fxn == "log", cond == "yes") |> 
  pull(model) |> 
  (\(x) x[[1]] )()

slopes <- coef(power_mod)$`subject:lab_id` |> 
  as_tibble(rownames = "subject")

slopes_cleaned <- slopes |> 
  mutate(subject = str_remove(subject, ":.*")) |> 
  select(subject, slope = trial_num)
```

```{r}
df_child <- d_proc |> 
  select(project_id, lab_id, unique_subject_identifier, method, age_days, participant_sex, 
         household_size, parent1_education, parent2_education, participant_place_of_birth,
         race, primarycaregiver_gender, language_group, primary_language, primary_language_exposure,
         caregiver_seat) |> 
  distinct() |> 
  left_join(slopes_cleaned, by = join_by(unique_subject_identifier == subject))

write_csv(df_child, here("data", "processed_data", "child_demogs_slopes.csv"))
```
