---
title: "MB Habituation - Habituation Curve"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(here)
library(zoo)

theme_set(theme_bw() + 
            theme(panel.grid = element_blank()))
```

## Load preprocessed data

```{r load}
d <- read_csv(here("data","processed_data","final_merged_manybabies_preprocessed.csv"))
  
d |> count(project_id, name = "n_trials")
```

```{r}
d_proc <- d |> 
  mutate(
    project_id = project_id |> factor() |> `contrasts<-`(value = contr.sum(3) * 0.5),
    looking_time_corr = ifelse(
      project_id == "MB4" & trial_type == "social",
      pmax(0.01, (looking_time - 4.4)),
      looking_time
    )) |> 
  group_by(unique_subject_identifier) |> 
  mutate(lt_mov_avg = rollmean(looking_time, 3, na.pad = T),
         lt_baseline = lt_mov_avg[2],
         lt_habituated = lt_mov_avg < 0.5 * lt_baseline)
```

```{r}
d_proc |> 
  ggplot(aes(x = trial_num, y = lt_mov_avg, col = project_id)) +
  geom_line(aes(group = unique_subject_identifier), alpha = .1)
```


## Models
```{r}
formula_list <- list(
  formula(looking_time ~ trial_num * project_id + (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ poly(trial_num, 2) * project_id + (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ log(trial_num) * project_id + (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ trial_num * project_id + (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ poly(trial_num, 2) * project_id + (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ log(trial_num) * project_id + (1 + trial_num | lab_id / subject))
)
```

```{r}
models <- tibble(
  formula = formula_list,
  model = map(formula_list, \(f) {
    lmer(f, data = d_proc)
  }),
  BIC = map_dbl(model, BIC))

models |> select(formula, BIC)
```

```{r}
d_preds <- d_proc |> 
  mutate(pred = predict(models |> slice_min(BIC) |> pull(model)))
```

Decision points
- Figure out MB4 looking time situation (DV: LT_video, LT_frame, LT_video_plus_frame)
- Figure out MB4 condition differences (do we correct for 4.4s extra in the social condition?)
- Moving average comparison
- Trial type interactions with trial_num 


```{r}
d_proc |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = looking_time_corr, fill = trial_type)) +
  geom_histogram()  
```

```{r}
d_preds |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = looking_time, y = lookingtime_freezeframe, col = lab_id, shape = trial_type)) +
  geom_hline(yintercept = 12) +
  geom_point(alpha = .3)
```

```{r}
lmer(
  log(looking_time_corr) ~ log(trial_num) + 
    (1 + trial_num | lab_id / subject),
  data = d_proc
) |> 
  summary()
```


```{r}
d_preds |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = trial_num, y = lookingtime_freezeframe, col = trial_type)) +
  geom_point(alpha = .2, position = position_jitter(width = .2)) +
  geom_smooth(alpha = .1)
```



## Linear functional form: LT ~ trial_num

### Raw looking time

```{r lin-raw}
m_lin_raw <- lmer(
  looking_time ~ trial_num * project_id + (1 + trial_num | lab_id / subject),
  data = d
)
summary(m_lin_raw)
```

### Log looking time (canonical form)

`log(LT) ~ trial_num` corresponds to exponential decay on the raw scale and is
the standard "linear" habituation specification.

```{r lin-log}
m_lin_log <- lmer(
  log_lt ~ trial_num + (1 + trial_num | subject) + (1 | project),
  data = d
)
summary(m_lin_log)
```

## Quadratic functional form: LT ~ poly(trial_num, 2)

Adds a squared term to capture curvature — e.g. faster decay early in the
sequence that flattens out later. `poly(trial_num, 2)` uses orthogonal
polynomials so the linear and quadratic terms are uncorrelated.

### Raw looking time

```{r quad-raw}
m_quad_raw <- lmer(
  looking_time ~ poly(trial_num, 2) + (1 + trial_num | subject) + (1 | project),
  data = d
)
summary(m_quad_raw)
```

### Log looking time

```{r quad-log}
m_quad_log <- lmer(
  log_lt ~ poly(trial_num, 2) + (1 + trial_num | subject) + (1 | project),
  data = d
)
summary(m_quad_log)
```

## Fixed-effect estimates (all models)

```{r tidy}
bind_rows(
  tidy(m_lin_raw,  effects = "fixed") |> mutate(model = "LT ~ trial_num"),
  tidy(m_lin_log,  effects = "fixed") |> mutate(model = "log(LT) ~ trial_num"),
  tidy(m_quad_raw, effects = "fixed") |> mutate(model = "LT ~ poly(trial_num, 2)"),
  tidy(m_quad_log, effects = "fixed") |> mutate(model = "log(LT) ~ poly(trial_num, 2)")
) |>
  select(model, term, estimate, std.error, statistic, p.value)
```

## Model comparison

```{r compare}
AIC(m_lin_raw, m_quad_raw)
AIC(m_lin_log, m_quad_log)
anova(m_lin_raw, m_quad_raw)
anova(m_lin_log, m_quad_log)
```

## Visual check

```{r plot, fig.width=8, fig.height=5}
d |>
  group_by(project_id, trial_num) |>
  summarise(mean_lt = mean(looking_time), .groups = "drop") |>
  ggplot(aes(trial_num, mean_lt, colour = project_id)) +
  geom_point() +
  geom_smooth(aes(linetype = "Linear"),    method = "lm",
              formula = y ~ x,             se = FALSE) +
  geom_smooth(aes(linetype = "Quadratic"), method = "lm",
              formula = y ~ poly(x, 2),    se = FALSE) +
  labs(x = "Trial number", y = "Mean looking time (s)",
       title = "Habituation: linear vs. quadratic", linetype = "Form") +
  theme_minimal()
```

```{r plot, fig.width=8, fig.height=5}
d |>
  group_by(project_id, trial_num) |>
  summarise(mean_log_lt = mean(log_lt), .groups = "drop") |>
  ggplot(aes(trial_num, mean_log_lt, colour = project_id)) +
  geom_point() +
  geom_smooth(aes(linetype = "Linear"),    method = "lm",
              formula = y ~ x,             se = FALSE) +
  geom_smooth(aes(linetype = "Quadratic"), method = "lm",
              formula = y ~ poly(x, 2),    se = FALSE) +
  labs(x = "Trial number", y = "Mean log looking time (s)",
       title = "Habituation: linear vs. quadratic", linetype = "Form") +
  theme_minimal()
```
