---
title: "MB Habituation - Habituation Curve"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(here)
```

## Load preprocessed data

```{r load}
d <- read_csv(here("data","processed_data","final_merged_manybabies_preprocessed.csv"))
d |> count(project_id, name = "n_trials")
```

## Linear functional form: LT ~ trial_num

### Raw looking time

```{r lin-raw}
m_lin_raw <- lmer(
  looking_time ~ trial_num + (1 + trial_num | subject) + (1 | project),
  data = d
)
summary(m_lin_raw)
```

### Log looking time (canonical form)

`log(LT) ~ trial_num` corresponds to exponential decay on the raw scale and is
the standard "linear" habituation specification.

```{r lin-log}
m_lin_log <- lmer(
  log_lt ~ trial_num + (1 + trial_num | subject) + (1 | project),
  data = d
)
summary(m_lin_log)
```

## Quadratic functional form: LT ~ poly(trial_num, 2)

Adds a squared term to capture curvature — e.g. faster decay early in the
sequence that flattens out later. `poly(trial_num, 2)` uses orthogonal
polynomials so the linear and quadratic terms are uncorrelated.

### Raw looking time

```{r quad-raw}
m_quad_raw <- lmer(
  looking_time ~ poly(trial_num, 2) + (1 + trial_num | subject) + (1 | project),
  data = d
)
summary(m_quad_raw)
```

### Log looking time

```{r quad-log}
m_quad_log <- lmer(
  log_lt ~ poly(trial_num, 2) + (1 + trial_num | subject) + (1 | project),
  data = d
)
summary(m_quad_log)
```

## Fixed-effect estimates (all models)

```{r tidy}
bind_rows(
  tidy(m_lin_raw,  effects = "fixed") |> mutate(model = "LT ~ trial_num"),
  tidy(m_lin_log,  effects = "fixed") |> mutate(model = "log(LT) ~ trial_num"),
  tidy(m_quad_raw, effects = "fixed") |> mutate(model = "LT ~ poly(trial_num, 2)"),
  tidy(m_quad_log, effects = "fixed") |> mutate(model = "log(LT) ~ poly(trial_num, 2)")
) |>
  select(model, term, estimate, std.error, statistic, p.value)
```

## Model comparison

```{r compare}
AIC(m_lin_raw, m_quad_raw)
AIC(m_lin_log, m_quad_log)
anova(m_lin_raw, m_quad_raw)
anova(m_lin_log, m_quad_log)
```

## Visual check

```{r plot, fig.width=8, fig.height=5}
d |>
  group_by(project_id, trial_num) |>
  summarise(mean_lt = mean(looking_time), .groups = "drop") |>
  ggplot(aes(trial_num, mean_lt, colour = project_id)) +
  geom_point() +
  geom_smooth(aes(linetype = "Linear"),    method = "lm",
              formula = y ~ x,             se = FALSE) +
  geom_smooth(aes(linetype = "Quadratic"), method = "lm",
              formula = y ~ poly(x, 2),    se = FALSE) +
  labs(x = "Trial number", y = "Mean looking time (s)",
       title = "Habituation: linear vs. quadratic", linetype = "Form") +
  theme_minimal()
```
