---
title: "MB Habituation - Habituation Curve"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(lmerTest)
library(janitor)
library(broom.mixed)
library(here)
library(zoo)

theme_set(theme_bw() + 
            theme(panel.grid = element_blank()))

FIRST_TIME <- TRUE
```

## Load preprocessed data

```{r load}
d <- read_csv(here("data","processed_data","final_merged_manybabies_preprocessed.csv"))
  
d |> count(project_id, name = "n_trials")
```

```{r}
d_proc <- d |> 
  mutate(
    project_id = project_id |> factor() |> `contrasts<-`(value = contr.sum(3) * 0.5),
    looking_time_corr = ifelse(
      project_id == "MB4" & trial_type == "social",
      pmax(0.01, (looking_time - 4.4)),
      looking_time
    )) |> 
  group_by(unique_subject_identifier) |> 
  mutate(lt_mov_avg = rollmean(looking_time, 3, na.pad = T),
         lt_baseline = lt_mov_avg[2],
         lt_habituated = lt_mov_avg < 0.5 * lt_baseline)
```

## Visual check

```{r plot 1, fig.width=8, fig.height=5}
d |>
  group_by(project_id, trial_num) |>
  summarise(mean_lt = mean(looking_time), .groups = "drop") |>
  ggplot(aes(trial_num, mean_lt, colour = project_id)) +
  geom_point() +
  geom_smooth(aes(linetype = "Linear"),    method = "lm",
              formula = y ~ x,             se = FALSE) +
  geom_smooth(aes(linetype = "Quadratic"), method = "lm",
              formula = y ~ poly(x, 2),    se = FALSE) +
  labs(x = "Trial number", y = "Mean looking time (s)",
       title = "Habituation: linear vs. quadratic", linetype = "Form") +
  theme_minimal()
```

```{r plot 2, fig.width=8, fig.height=5}
d |>
  group_by(project_id, trial_num) |>
  summarise(mean_log_lt = mean(log_lt), .groups = "drop") |>
  ggplot(aes(trial_num, mean_log_lt, colour = project_id)) +
  geom_point() +
  geom_smooth(aes(linetype = "Linear"),    method = "lm",
              formula = y ~ x,             se = FALSE) +
  geom_smooth(aes(linetype = "Quadratic"), method = "lm",
              formula = y ~ poly(x, 2),    se = FALSE) +
  labs(x = "Trial number", y = "Mean log looking time (s)",
       title = "Habituation: linear vs. quadratic", linetype = "Form") +
  theme_minimal()
```

## MB4 DV exploration
MB4 has two DVs: the LT on the video itself, and the LT on the freeze frame at the end of the video. 
Furthermore, the videos differed across conditions, with the social condition video being 4.4s longer than the nonsocial condition video. 

We explore the goodness of different DV choices (LT_video, LT_frame, LT_video_frame, LT_video_corr, LT_video_frame_corr).
One option is to fit the power law fit and see which one has the best fit.

```{r eval=FIRST_TIME}
mb4_formula_list <- list(
  formula(log(looking_time + .01) ~ log(trial_num) * trial_type + 
            (1 + log(trial_num) | lab_id / subject)),
  formula(log(lookingtime_freezeframe + .01) ~ log(trial_num) * trial_type + 
            (1 + log(trial_num) | lab_id / subject)),
  formula(log(looking_time + lookingtime_freezeframe + .01) ~ log(trial_num) * trial_type + 
            (1 + log(trial_num) | lab_id / subject)),
  formula(log(looking_time_corr + .01) ~ log(trial_num) * trial_type + 
            (1 + log(trial_num) | lab_id / subject)),
  formula(log(looking_time_corr + lookingtime_freezeframe + .01) ~ log(trial_num) * trial_type + 
            (1 + log(trial_num) | lab_id / subject))
)

mb4_models <- tibble(
  formula = mb4_formula_list,
  model = map(formula, \(f) {
    lmer(f, data = d_proc)
  }),
  BIC = map_dbl(model, BIC)
)

mb4_models |> select(-model)

saveRDS(mb4_models, here("models", "mb4_models.rds"))
```

Observations:
- Models with cumulative LT do not converge
- The best fitting model was the model with LT_frame.
We switch to LT_frame for all subsequent analyses.

```{r}
d_proc <- d_proc |> 
  mutate(
    lookingtime_video = ifelse(
      project_id == "MB4",
      looking_time,
      NA
    ),
    looking_time = ifelse(
      project_id == "MB4",
      lookingtime_freezeframe + .001,
      looking_time
    ))

d_proc <- d_proc |> 
  group_by(unique_subject_identifier) |> 
  mutate(lt_mov_avg = rollmean(looking_time, 3, na.pad = T),
         lt_baseline = lt_mov_avg[2],
         lt_habituated = lt_mov_avg < 0.5 * lt_baseline) |> 
  ungroup()

write_csv(d_proc, here("data", "processed_data", "reprocessed_data.csv"))
```

## Moving average habituation criterion

```{r}
d_habituated <- d_proc |> 
  filter(lt_habituated) |> 
  group_by(unique_subject_identifier) |> 
  slice(1)

d_habituated |> 
  ggplot(aes(x = trial_num, fill = project_id)) +
  geom_histogram(binwidth = 1)
```

## Models
```{r}
formula_list <- list(
  formula(looking_time ~ trial_num * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ poly(trial_num, 2) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ log(trial_num) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ trial_num * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ poly(trial_num, 2) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ log(trial_num) * project_id + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ trial_num * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ poly(trial_num, 2) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(looking_time ~ log(trial_num) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ trial_num * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ poly(trial_num, 2) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  formula(log(looking_time) ~ log(trial_num) * project_id * trial_type + 
            (1 + trial_num | lab_id / subject)),
  
  formula(looking_time ~ poly(trial_num, 2) * project_id + 
            (1 + poly(trial_num, 2) | lab_id / subject)),
  formula(looking_time ~ log(trial_num) * project_id + 
            (1 + log(trial_num) | lab_id / subject)),
  formula(log(looking_time) ~ poly(trial_num, 2) * project_id + 
            (1 + poly(trial_num, 2) | lab_id / subject)),
  formula(log(looking_time) ~ log(trial_num) * project_id + 
            (1 + log(trial_num) | lab_id / subject)),
  formula(looking_time ~ poly(trial_num, 2) * project_id * trial_type + 
            (1 + poly(trial_num, 2) | lab_id / subject)),
  formula(looking_time ~ log(trial_num) * project_id * trial_type + 
            (1 + log(trial_num) | lab_id / subject)),
  formula(log(looking_time) ~ poly(trial_num, 2) * project_id * trial_type + 
            (1 + poly(trial_num, 2) | lab_id / subject)),
  formula(log(looking_time) ~ log(trial_num) * project_id * trial_type + 
            (1 + log(trial_num) | lab_id / subject))
)
```

```{r eval=FIRST_TIME}
models <- tibble(
  lt = c(rep(c("lt", "log_lt"), times = 2, each = 3), 
         rep(c("lt", "log_lt"), times = 2, each = 2)),
  fxn = c(rep(c("linear", "quadratic", "log"), times = 4),
          rep(c("quadratic", "log"), times = 4)),
  cond = c(rep(c("no", "yes"), each = 6),
           rep(c("no", "yes"), each = 4)),
  full_re = c(rep("no", each = 12), rep("yes", each = 8)),
  formula = formula_list,
  model = map(formula, \(f) {
    lmer(f, data = d_proc)
  }),
  BIC = map_dbl(model, BIC)
)

saveRDS(models, here("models", "fit_models.rds"))
```

```{r}
models <- readRDS(here("models", "fit_models.rds"))

models |> select(-formula, -model) |> arrange(BIC)

models |> 
  ggplot(aes(x = BIC, y = interaction(fxn, full_re), col = cond)) +
  geom_point() +
  facet_grid(~ lt, scales = "free_x")
```

# Expanding window CV
```{r eval=FIRST_TIME}
models_cv <- models |> 
  filter(cond == "yes",
         # fxn == "linear", # for testing
         lt == "log_lt") |> 
  select(-model, -BIC) |> 
  cross_join(tibble(last_trial = 4:15)) |> 
  mutate(model = map2(formula, last_trial, \(f, l) {
    lmer(f, data = d_proc |> filter(trial_num <= l))
  }),
  BIC = map_dbl(model, BIC)
  )

models_cv_preds <- models_cv |> 
  mutate(d_filtered = map(last_trial, \(l) {
    d_proc |> 
      filter(trial_num == (l + 1),
             subject %in% (d_proc |> 
                             filter(trial_num <= l) |> 
                             pull(subject) |> 
                             unique())) |> 
      ungroup()
  }),
  preds = map2(model, d_filtered, \(m, d) {
    d |> 
      mutate(pred = predict(m, d))
  }), rmse = map_dbl(preds, \(p) {
    p |> summarise(rmse = sqrt(sum((pred - log_lt) ^ 2) / n())) |> 
      pull(rmse)
  }))

saveRDS(models_cv_preds, here("models", "fit_cv_preds.rds"))
```

```{r eval=FIRST_TIME}
models_cv_ci <- models_cv_preds |> 
  mutate(rmse = map(preds, \(p) {
    p_sq <- p |> mutate(sqerr = (pred - log_lt) ^ 2)
    boot::boot(p_sq, \(d, i) {
      d[i,] |> pull(sqerr) |> mean() |> sqrt()
    }, 1000)
  })) |> 
  select(lt, fxn, cond, full_re, last_trial, rmse, preds)

models_cv_ci <- models_cv_ci |> 
  mutate(ci = map(rmse, boot::boot.ci),
         mean = map_dbl(preds, \(p) {
           p |> summarise(rmse = sqrt(sum((pred - log_lt) ^ 2) / n())) |> 
             pull(rmse)
         }))

saveRDS(models_cv_ci, here("models", "fit_cv_ci.rds"))
```


```{r}
models_cv_preds <- readRDS(here("models", "fit_cv_preds.rds"))
models_cv_ci <- readRDS(here("models", "fit_cv_ci.rds"))

models_cv_ci_simp <- models_cv_ci |> 
  mutate(ci = map(ci, \(c) {
    c$normal |> as.tibble() |> `colnames<-`(c("ci_size", "ci.lb", "ci.ub"))
  }) |> list_rbind()) |> 
  unnest(ci) |> 
  select(-rmse, -preds)
```

```{r}
models_cv_preds |> 
  select(-preds) |> 
  ggplot(aes(x = last_trial, y = rmse, col = interaction(lt, fxn, cond, full_re))) +
  geom_errorbar(aes(ymin = ci.lb, ymax = ci.ub, y = mean),
                data = models_cv_ci_simp, width = 0,
                position = position_dodge(.4)) +
  geom_point(position = position_dodge(.4)) +
  geom_line(aes(group = interaction(lt, fxn, cond, full_re)), position = position_dodge(.4)) +
  labs(x = "Last trial observed", y = "Prediction RMSE on next trial")
```

```{r}
models_cv_preds |> 
  select(-preds) |> 
  group_by(last_trial) |> 
  mutate(rmse_diff = rmse - min(rmse)) |> 
  ggplot(aes(x = last_trial, y = rmse_diff, col = interaction(lt, fxn, cond, full_re))) +
  # geom_errorbar(aes(ymin = ci.lb, ymax = ci.ub, y = mean),
  #               data = models_cv_ci_simp, width = 0,
  #               position = position_dodge(.4)) +
  geom_point(position = position_dodge(.4)) +
  geom_line(aes(group = interaction(lt, fxn, cond, full_re)), position = position_dodge(.4)) +
  labs(x = "Last trial observed", y = "ΔNext-trial prediction RMSE from best")
```

```{r}
models_cv_preds |> 
  select(-preds) |> 
  group_by(last_trial) |> 
  mutate(bic_diff = BIC - min(BIC)) |> 
  ggplot(aes(x = last_trial, y = bic_diff, col = interaction(lt, fxn, cond, full_re))) +
  geom_point(position = position_dodge(.4)) +
  geom_line(aes(group = interaction(lt, fxn, cond, full_re)), position = position_dodge(.4)) +
  labs(x = "Last trial observed", y = "ΔBIC from best")
```

```{r}
d_preds <- d_proc |> 
  filter(!is.na(looking_time)) |> 
  ungroup() |> 
  mutate(pred = predict(models$model[[12]]))
```

## Some plots

```{r}
d_proc |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = looking_time_corr, fill = trial_type)) +
  geom_histogram()  
```

```{r}
d_preds |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = looking_time, y = lookingtime_freezeframe, col = lab_id, shape = trial_type)) +
  geom_hline(yintercept = 12) +
  geom_point(alpha = .3)
```

```{r}
d_preds |> 
  filter(project_id == "MB4") |> 
  ggplot(aes(x = trial_num, y = lookingtime_freezeframe, col = trial_type)) +
  geom_point(alpha = .2, position = position_jitter(width = .2)) +
  geom_smooth(alpha = .1)
```

```{r}
d_proc |> 
  filter(project_id == "MB1",
         !is.na(participant_sex)) |> 
  ggplot(aes(x = trial_num, y = looking_time, 
             col = cut(age_days, 
                       breaks = quantile(age_days, probs = seq(0, 1, 0.25)), 
                       include.lowest = TRUE))) +
  geom_point(alpha = .2, position = position_jitter(width = .2)) +
  geom_smooth(alpha = .3) +
  theme(legend.position = "bottom") +
  labs(col = "Covariate")
```

## Quantifying each child's slope
```{r}
power_mod <- models |> 
  filter(lt == "log_lt", fxn == "log", cond == "yes", full_re == "yes") |> 
  pull(model) |> 
  (\(x) x[[1]])()

slopes <- coef(power_mod)$`subject:lab_id` |> 
  as_tibble(rownames = "subject")

slopes_cleaned <- slopes |> 
  mutate(subject = str_remove(subject, ":.*")) |> 
  select(subject, slope = `log(trial_num)`, intercept = `(Intercept)`)
```

```{r}
lab_id <- read_csv(here("data", "raw_data", "mb_labid.csv")) |> 
  clean_names() |> 
  mutate(continent = recode_values(
    country,
    c("United States", "Canada", "México", "Colombia") ~ "Americas",
    c("Austria", "Netherlands", "United Kingdom", "Italy",
      "Belgium", "Norway", "Hungary", "Germany", "Czechia", 
      "Sweden", "Spain", "France", "Poland", "Iceland",
      "Switzerland") ~ "Europe",
    c("South Korea", "Japan", "Israel", "Singapore",
      "Turkey", "Hong Kong", "China") ~ "Asia",
    c("Australia", "New Zealand") ~ "Oceania"
  ))
```


```{r}
df_child <- d_proc |> 
  select(project_id, lab_id, unique_subject_identifier, method, age_days, participant_sex, 
         household_size, parent1_education, parent2_education, participant_place_of_birth,
         race, primarycaregiver_gender, language_group, primary_language, primary_language_exposure,
         caregiver_seat) |> 
  distinct() |> 
  left_join(slopes_cleaned, by = join_by(unique_subject_identifier == subject)) |> 
  left_join(d_habituated |> select(unique_subject_identifier, hab_trial = trial_num),
            by = join_by(unique_subject_identifier)) |> 
  mutate(lab_id_l = str_to_lower(lab_id) |> 
           replace_values(
             "minddevlabbicocca" ~ "biccbabylab",
             "pocdnorthwestern" ~ "icdcnorthwestern",
             "babylab_leiden" ~ "babylableiden",
             "weltentdecker-zurich" ~ "weltentdeckerzurich",
             "cogcienmat_univalle" ~ "cogcienmatunivalle",
             "gauggottingen" ~ "gauggöttingen",
             "ikidstutsc" ~ "ikidsutsc",
             "infantcog_ubc" ~ "infantcogubc",
             "princetonbabylab" ~ "babylabprinceton"
           )) |> 
  left_join(lab_id |> select(lab_id, continent) |> mutate(lab_id = str_to_lower(lab_id)),
            by = join_by(lab_id_l == lab_id)) |> 
  select(-lab_id_l)

write_csv(df_child, here("data", "processed_data", "child_demogs_slopes.csv"))
```

```{r}
df_child |> 
  ggplot(aes(x = age_days, y = intercept, col = project_id)) +
  geom_point(alpha = .2) +
  geom_hline(yintercept = 2, lty = "dashed") +
  geom_smooth(method = "lm")
```

```{r}
df_child |> 
  ggplot(aes(x = hab_trial, y = slope, col = project_id)) +
  geom_point(alpha = .2, position = position_jitter(width = .4)) +
  geom_smooth(method = "lm", se = F) +
  labs(x = "Trial when moving average habituation criterion reached",
       y = "Participant LT slope",
       col = "Project")
```


